# HYPOTHESIS TESTING 
## Introduction

Hypothesis testing is a standard statistical procedure for deciding between two competing alternatives or scenarios, based on evidence presented by a data sample collected for the purpose of conducting the test. Most descriptive statistics such as mean, median and standard deviation are usually computed from data samples which may vary with the whole population. For instance, we want to find the concentration of hexavalent chromium, a human carcinogen, in soil. We will collect random 20 samples. The average concentration of the carcinogen from the collected samples will at most vary with the average concentration in the entire land, that's where **hypothesis testing** comes in. 

**Basic terminology and procedure for hypothesis testing**

There is a need to understand the basic terms before diving deep into the process of hypothesis testing. If we have historical data, the presence of a trend must be established and below are the assumptions that should be made prior to conducting hypothesis testing; 

i. The data is assumed to belong to a specified distribution(typically normal distribution). 
ii. The data has a stable or stationary trend(free from of temporal or spatial trend).
iii. There are no unusual observations or outliers. 

The hypothesis testing that is based on the above is described as **parametric hypothesis testing**. Otherwise, if the above assumptions are not met, the test is known as **non-parametric hypothesis testing**. 

* **Null hypothesis** ($H_0$): Is a statement of the presumed default or baseline condition. In this case the null hypothesis is that there is no difference between the carcinogenic concentration is below the allowed limit. 

* **Alternate hypothesis** ($H_1$ or $H_A$): is the inverse or the opposite of the null hypothesis. Alternate hypothesis contradicts the statement of the presumed default. In this case, the alternate hypothesis states that the carcinogenic level is above the allowed limit.

The evidence produced by the sample after conducting hypothesis testing can either refute or support the null hypothesis.

* The **one-tailed (One-Sided) Hypothesis** test checks for an effect in only one direction. In our case, we test if the carcinogen levels exceed the allowed limit(right side). Alternatively, it can be checked of it is below the allowed limits (left-sided). 

* **Upper-Tailed Hypothesis Test** is the same as the right one-tailed hypothesis test that checks if the value of interest exceed the null hypothesis value

* **Lower-Tailed Hypothesis Test** is the left one-tailed hypthesis test where it checks if the value of interest is below the null hypothesis value. In our case the allowed limit for carcinogens in soil. 

* **Two-Tailed Hypothesis Test** checks for the effect in both directions. In our we can change to find if the carcinogen levels are below or above the allowed limit. 

* **T-Value** is the test statistic calculated from the sample data to determine how far the sample mean is from the hypothesized mean, in terms of standard error. 

* **Critical T-Value** is the threshold value for the t-distribution that corresponds to the significance level. If the t-value exceeds the critical t-value, the null hypothesis is rejected. 

* **Significant level**($\alpha$) is the probability threshold for rejecting the null hypothesis, usually set to 0.05(5%). 

In most cases if the probability of observing the test statistic (p-value) is less than 0.005, the null hypothesis is rejected. 

* **Degrees of Freedom(df)** are the number of independent values in the data sample that are free to vary when calculating a statistic. 

In most cases **df** is calculated as $$df = n - 1$$

where $n$ represents the sample size(count). 

In our cases where we have 20 samples therefore $df = 20 -1$ and the degree of freedom will be 19

* **P-Value** indicates the probability of obtaining a test statistic at least as extreme as the observed one, assuming the null hypothesis is true.

Below are the steps taken in hypothesis testing; 

i. Formulate the null and alternate hypothesis. 
ii. Assume the null hypothesis is true and compute a test statistic that follows a known distribution. 
iii. Calculate a test statistic value based on the data sample which follows a known distribution and determine the critical value for chosen distribution at a given significance level. 
iv. Compare the test statistic to the critical value to assess the plausibility of the null hypothesis. For symettrical distributions(e.g t, normal), critical values are located on;

  - *upper-tailed test*: positive side of the distribution 
  - *lower-tailed test*: negative side of the distribution
  - *two-tailed test*: both sides, with equal magnitude

v. Calculate the p-value to find how the probability of obtaining a test statistic more extreme than the observed value. 

<span style="color: orange;">**Try it!**</span> 

To get the steps above clear, lets work on an example. This is an hypothetical problem where a regulatory body sets that the acceptable phosphorous concentration in a lake at **1.5mg/L**. 30 sample measurements were collected at different locations in the lake. The objective was to determine if the phosphorous levels are significantly higher than the allowed limit. 

Lets simulate the data. 
```{r}
set.seed(123) # for reproducibility

phosphorous_levels <- rnorm(n = 30,
                            mean = 1.7,
                            sd = 0.2)
```

* Formulate the null hypothesis 

- Null Hypothesis($H_0$): $\mu$ < 1.5 - phosphorous levels are within the acceptable levels.
- Alternate Hypothesis ($H_1$): $\mu$ > 1.5 - phosphorous levels exceed the acceptable limit. 

This is a right-sided(one-sided) test since we are only interested in detecting an increase in phosphorous levels

* We use a one-sample ttest since the population data standard deviation is unknown, and the sample size is small(n=30). The formula for t-statistic is calculated as; $$t = {{\overline{x} - \mu_0}\over{s/\sqrt{n}}}$$
Where: 

- $\overline{x}$ is the sample mean.
- $\mu_0$ is the population mean under null hypothesis (1.5mg/L)
- $s$ is the sample standard deviation.
- $n$ is the sample size. 

* Perform the one-sample t-test
```{r}
# One sample t-test 
t_test_result <- t.test(phosphorous_levels, 
                        mu = 1.5,
                        alternative = "greater")

# Display the results 
print(t_test_result)
```

* Determine the critical value

- For the right-tailed test at a 5% significance levels($alpha$=0.05), the critical value is the t-value that corresponds to the top 5% of the t-distribution. The `qt()` function will be used to calculate this; $$t_{critical} = qt(1-\alpha, df = n - 1)$$

```{r}
critical_t <- qt(0.95, df = length(phosphorous_levels) - 1)

critical_t
```

The t-statistic  is greater than the critical t therefore the null hypothesis is rejected. Lets confirm this by also computing the p_value

* Calculate the null hypothesis 
```{r}
p_value <- t_test_result$p.value
p_value
```

The null hypothesis is rejected as the p_value is less than 0.05. Therefore, we make conclusion that the phosphorous levels in the lake exceed the acceptable levels. 

## Parametric tests 
As described before parametric tests rely on the assumptions that the data follows a known distribution (commonly the normal distribution) and it should be free of major outliers. In case where the test requires the data to follow a normal distribution, if the sample size is small it must follow a normal distribution while for larger sample size (e.g n≥30) the **Central Limit Theorem** allows the researchers to approximate normality even of the data is not perfectly normal

### Parametric single-sample test 

This type of test is used to determine whether the mean of a single sample is significantly different from a known or hypothesized population mean. It assumes that the sample data comes from a population that follows a normal distribution (or has a large enough sample size to approximate normality).

The **main objective** of this test is to *test whether the sample mean differs significantly from a reference or target value, such as an industry standard, a population average, or a theoretical expectation in hypothesis testing*

These are the steps taken for One-Sample T-test;

1. *State the Hypotheses:*

- **Null Hypothesis**($H_0$): The sample mean equals the population mean ($\mu = \mu_0$)
- **Alternate Hypothesis**($H_A$): The sample mean is different from the population mean($\mu \neq \mu_0$)

2. *Compute the test statistic*

The formula for test statistic is $$t = {{\overline x - \mu_0}\over{s/\sqrt n}}$$

Where: 

- $\overline x$: is the sample mean
- $\mu_0$: is the hypothesized population mean 
- $s$: is the sample standard deviation
- $n$: is the sample size 

3. Determine the degree of freedom as $df = n- 1$
4. **Compare $t$ to critical value or use p-value**

### Parametric two-sample test

## Nonparametric test
### Nonparametric two-sample paired sign test and paired Wilcoxon Signed Rank Test 

## One-Way ANOVA 
### Parametric One-Way ANOVA 

### Nonparametric One-Way ANOVA (Kruskal-Wallis Test)


## <span style="color: green;">**Hands-on Exercise**</span> 

_______________________________________________________________________
<span style="color: brown;">**Solution**</span> 

<span style="color: brown;">**________________________________________________________________________________**</span>